---
title: "San Francisco Crime Visualization"
author: "ww44ss"
date: "Oct  2014"
output: pdf_document

---

##Synopisis

This analysis looks at San Francisco crime data. It addresses the questions:  
- Does the number of crimes show obvious day to day variation, particularly between weekdays and weekends?    
- How does crime vary day to day on a per district basis?  
- Can we see hotspots where crimes are most prevalent in San Francisco?
- Do particular crimes happen more frequenly at different times of day?

This data suggest that police patrols can be optimized to specific districts and locations to focus on particular crimes. 


##Get Data


```{r "data file information", echo=FALSE, message=FALSE, warning=FALSE}

datadir <-"sfpd"
files <- list.files(datadir)
len<-length(files)

```

There were `r len` files found in the data directory `r getwd()`.  

Data cleaning was pretty straight forward. Set factors to makes days of week follow standard order (instead of default alphabetical). Convert Date to a r date format. Time I just chose to bucket by hour rather than convert to hh:mm format, which was too fine grained. 

```{r "get the data", echo=FALSE, message=FALSE, warning=FALSE}

datafile = "SFPD_Incidents_-_Previous_Three_Months.csv"
file = paste0(datadir,"/",datafile)    

data <- read.csv(file)

        ##fix day of week order
        data$DayOfWeek <- factor(data$DayOfWeek, levels= c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

        ##for simplicity keep only complete cases
        data<-data[complete.cases(data),]

        data$Date<-as.Date(as.character(data$Date), "%m/%d/%Y")

        #had to be a little clever with the time , basically trick everything to think it is the same day.
        #data$Time<-as.POSIXct(paste("1970-01-01", as.character(data$Time)), format="%Y-%m-%d %H:%M")

        ##while the aboce works. I found the analysis looks easier just hacking out the hour....
        ##Alternate method is just to that the hh and convert to a number
        data$Time <- as.numeric(substring(as.character(data$Time),1,2))

str(data)

        ddata<-dim(data)


```

The above shows the structure of the data. There are statistics on `r ddata[1]` crimes in the file `r "datafile"`. 

##Analysis


####Question 1: How does crime vary day to day on a per district basis?  

```{r "plot of data2", fig.height=4, fig.width=7,  echo=FALSE, message=FALSE, warning=FALSE, results="asis"}

require(ggplot2)
require(plyr)

dataX <- ddply(data,.(DayOfWeek, PdDistrict), nrow)

        ##plot the data

        plotA <- ggplot(dataX, aes(x= factor(DayOfWeek), y = V1))
        plotA<- plotA+ geom_point(colour="blue")
        plotA<-plotA+facet_wrap(~PdDistrict, ncol=5) 
        plotA<-plotA+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
        #plotA<-plotA+coord_flip()

        print(plotA)


```

To analyze ply the data apart by day and district. At this point the analysis looks at total crime reports only. Later I'll look at types of crimes. 

District by district crime rates show variation in the day of the week. Each district has some pretty unique variation.  Some of the more interesting ones are listed below.
- _Southern_ has the highest crime rate by far, strongly peaking on Friday night.  
- _Bayview_ is mostly flat, but seems to show a higher rate on Friday nights.   
- _Central_ shows a strong upward trend on the weekends, with Friday and Saturday night showing about 20% increase in crime.   
- _Mission_ while having an overall farily high crime rate, shows little variation.   
- _Tenderloin_ shows an apparent drop in the crime rate.  

####Question 2: Does the leading type of crime vary by district?  

Observing the variability of crime by district its natural to ask whether the nature of crimes show any district by district distinction. The easiest way to get at this is to just pull the data aprt by district and sort. First let's just look citywide.  

```{r "most prevalent crimes", echo=FALSE, message=FALSE, warning=FALSE}

## print an html table of the most prevatlent crimes
        library(xtable)

        ## sort the data
        SF <- sort(table(data$Category), decreasing=TRUE)
        ## turn table into data frame
        SF<-as.data.frame(SF)
        ## just dump the first six lines
        head(SF, n=6)

```

By district the results show some variation. 

```{r "most prevalent crimes by district", echo=FALSE, message=FALSE, warning=FALSE}

## print an html table of the most prevatlent crimes

        ##subset the data for a few specific districts
        PlotTenderloin <-data[data$PdDistrict=="TENDERLOIN", ]
        PlotMission<-data[data$PdDistrict=="MISSION", ]
        PlotNorthern<-data[data$PdDistrict=="NORTHERN", ]   
        PlotRichmond<-data[data$PdDistrict=="RICHMOND", ]  


        ##Mission District
        ## sort the data
        ctable <- sort(table(PlotMission$Category), decreasing=TRUE)
        ctable<-as.data.frame(ctable)
        ## keep only top ten
        print("MISSION")
        head(ctable, n=5)

        ##Data for Richmond
         ## sort the data
        ctable <- sort(table(PlotRichmond$Category), decreasing=TRUE)
        ctable<-as.data.frame(ctable)
        ## keep only top ten
        print("RICHMOND")
        head(ctable, n=5)
        


```


This starts to show some of the richness of the data. For instance in the _Mission District_ while Larceny/Theft is the most prevalent item, assualt and drugs/narcotic violations together account for more total crime than the does Larceny/Theft.  
In the _Richmond District_, by contrast, Assault is not among the top six items, while vandalism and vehicle theft together account for less than half of the leading crime, again Larceny/Theft.  

Hence, although the leading type of crime does not vary by district, the top crimes shows marked variation depending on the district.  

####Question 3: Are there crime hotspots?

Here the hypothesis is there are "hot spots" where specific crimes tend to be localized. We can answer this by plotting crime types geographically. The easiest way to see this is to map the results. To speed up analysis I've chosen to focus only an a few "top" crimes from the lists above. Namely Larceny/Theft, Vehicle Theft,  and Assault.

```{r map_it, fig.height=9, fig.width=9, echo=FALSE, message=FALSE, warning=FALSE}
require(ggmap)
require(mapproj)

PlotTheft <-data[data$Category=="LARCENY/THEFT", ]
PlotVehicle<-data[data$Category=="VEHICLE THEFT", ]
PlotAssault<-data[data$Category=="ASSAULT", ] 
PlotVandalism<-data[data$Category=="VANDALISM", ] 

        ##get map data
        map <- get_map(source="google", maptype="roadmap", location = 'San Francisco', zoom = 13)
        ##generate map
        map1 <- ggmap(map) 

        ## 
        map1<-map1 + geom_point(aes(x = PlotTheft$X, y = PlotTheft$Y), data = PlotTheft, alpha = .1, color="red", size = 3)
        map1<-map1 + geom_point(aes(x = PlotAssault$X, y = PlotAssault$Y), data = PlotAssault, alpha = .1, color="blue", size = 3)
        map1<-map1 + geom_point(aes(x = PlotVehicle$X, y = PlotVehicle$Y), data = PlotVehicle, alpha = .2, color="darkgreen", size = 3)
        ##Supress Vandalism as it is just TMI.
        #map1<-map1 + geom_point(aes(x = PlotVandalism$X, y = PlotVandalism$Y), data = PlotVandalism, alpha = .2, color="orange", size = 3)
        map1


```

Clear hotspots are visible

The Map shows locations of crimes,  
_red_ data points correpond to thefts: these appear to be loaclized to mainly tourist areas.  
_blue_ data points representing Assault appear localized in the Tenderloin, Mission, adn Broadway areas.  
_DarkGreen_ data points representing Vehicle Theft are more spread across the City but appear most prevalent in residential areas.

####Question 4: Do crimes have unique time correlation signatures?  
  
Let's first look at Larcency data: 
```{r "by_time_of_day", fig.height=3, fig.width=5, echo=FALSE, message=FALSE, warning=FALSE}

        ##ddply the Larceny time of day
        TheftTime<-ddply(PlotTheft, .(Time), nrow)

        plot1<-ggplot(TheftTime, aes(x=Time, y=V1))+geom_point(colour="red", size=2)+geom_smooth()
        plot1<-plot1+xlab("Time")+ylab("Number")+ggtitle("Theft/Larcency v. Time")
        plot1<-plot1+ylim(c(0, 850))
        
        plot1
```    

This looks very different from the Assault data below.  

```{r "time2" , fig.height=3, fig.width=5, echo=FALSE, message=FALSE, warning=FALSE }

        ##ddply the Assault time of day
        AssaultTime<-ddply(PlotAssault, .(Time), nrow)

        plot1<-ggplot(AssaultTime, aes(x=Time, y=V1))+geom_point(colour="blue", size=2)+geom_smooth()
        plot1<-plot1+xlab("Time")+ylab("Number")+ggtitle("Assault v. Time")
        plot1<-plot1+ylim(c(0, 200))
        
        plot1

```

And vehicle threft shows an even more pronounced behavior.  

```{r "time3" , fig.height=3, fig.width=5, echo=FALSE, message=FALSE, warning=FALSE }
        ##ddply the Vehicle Theft time of day
        VehicleTime<-ddply(PlotVehicle, .(Time), nrow)

        plot1<-ggplot(VehicleTime, aes(x=Time, y=V1))+geom_point(colour="darkgreen", size=2)+geom_smooth()
        plot1<-plot1+xlab("Time")+ylab("Number")+ggtitle("Vehicle Theft v. Time")
        plot1<-plot1+ylim(c(0, 200))
        
        plot1



```

Crimes seem to show distinct time behavior. For instance Theft and Larceny appear to be low during morning hours, but peak around 6 pm.  Vehicle theft , on the other hand, picks up only after about 6 pm and drops off after midnight.

##Conclusions

This quick exploratory analysis found that crime frequency and type vary strongly by location in the city and also by time of day. Taking the data at face value, it suggests that plic patrols could be optimized for time and location, especially when targeting specific crimes.

There is some interesting analysis that could be done as a follow-up. For instance looking deeper at the time/location correlation of specific crimes. This date could be used to test the effectiveness of particular patrol and enforcement strategies.    
   
   
    
    
       
          